---
title: "Intro to R"
subtitle: "CSAS 2026"
author: "Tyler Adams"
format:
    revealjs:
        slide-number: true
        preview-links: true
        theme: default
---

## Prerequisites

- A laptop with [R](https://cran.r-project.org/)
and [Positron](https://positron.posit.co/download.html)
/[RStudio](https://posit.co/download/rstudio-desktop/) installed

- No experience needed

## Outline

- R basics

- R Syntax

- Data structure

- Practical R uses

## Why R?

- Runs on all major operating systems

- Open source

- Intuitive syntax

- User-created packages constantly provide new features

## Positron and RStudio

- Both Integrated Development Environments (IDEs) created
by posit

- RSudio is older, Positron released very recently

- Both have 4 main sections
  - Code editor
  - Variables
  - Console
  - Plots

## R syntax

- Type code into console

```{r}
#| echo: true
#| eval: false
# Example 1
print("Hello world")
```
```{r}
#| echo: true
#| eval: false
# Example 2
x <- 2
y = 3
```

- Both "=" and "<-" can be used to assign values to variables

- Look at the variables window

## R syntax, continued

- Basic operation signs

```{r}
#| echo: true
#| eval: false
# Example 3
# Addition
print(2 + 2)

# Subtraction
print(4 - 2)

# Multiplication (variables can be used too)
print(y * x)

# Division
print(x / 3)

# Exponentiation
print(x^2)
# or
print(x**2)
```

## Data Structure

- Real data mostly consists of 4 types

  - Character
  
  - Logical
  
  - Integer
  
  - Numeric
  
  
## Character Data

- Text information

  - Called strings in Python

- Contained in "" or ''

```{r}
#| echo: true
#| eval: false
# Example 4

fav_color <- "black"

print(fav_color)

```


## Logical data

- Also know as binary data, either TRUE or FALSE

- TRUE and FALSE are not contained within quotes

```{r}
#| echo: true
#| eval: false

# Example 5
x <- TRUE

print(x)


# Used for conditionals

if (x) {
  print("Yes")
}
```

## Integer and Numeric Data

- Numeric data is all real numbers, can be decimal or not

- Integers are whole real numbers

```{r}
#| echo: true
#| eval: false
# Example 6

x <- 3.4
y <- 4
z <- 4L

print(class(x))
print(class(y))
print(class(z))

x <- as.integer(x)

print(class(x))
print(x)
```

## Local Data

- Data can be imputed directly into R

- Tables can be useful for displaying data

```{r}
#| echo: true
#| eval: false
#Example 7

# Creates the numerical data in the table
tab <- as.table(rbind(c(762, 327, 468, 593), c(484, 239, 477, 5930)))

# Labels the rows then the columns 
dimnames(tab) <- list(c("Honors", "Not Honors"), Grade = c("9th", "10th", "11th", "12th"))

```

