---
title: "Introduction to R"
subtitle: "CSAS 2026"
author: "Tyler Adams"
format:
    revealjs:
        slide-number: true
        preview-links: true
        theme: default
---

## Prerequisites

-   A laptop with [R](https://cran.r-project.org/) and [Positron](https://positron.posit.co/download.html) /[RStudio](https://posit.co/download/rstudio-desktop/) installed

-   No experience needed

## Outline

-   R basics

-   R Syntax

-   Data structure

-   Practical R uses

## Why R?

-   Runs on all major operating systems

-   Open source

-   Intuitive syntax

-   User-created packages constantly provide new features

## Positron and RStudio

-   Both Integrated Development Environments (IDEs) created by posit

-   RSudio is older, Positron released very recently

-   Both have 4 main sections

    -   Code editor
    -   Variables
    -   Console
    -   Plots

## R syntax

-   Type code into console

```{r}
#| echo: true
#| eval: false
print("Hello world")
```

```{r}
#| echo: true
#| eval: false
x <- 2
y = 3
```

-   Both "=" and "\<-" can be used to assign values to variables

-   Look at the variables window

## R syntax, continued

-   Basic operation signs

```{r}
#| echo: true
#| eval: false

# Addition
print(2 + 2)

# Subtraction
print(4 - 2)

# Multiplication (variables can be used too)
print(y * x)

# Division
print(x / 3)

# Exponentiation
print(x^2)
# or
print(x**2)
```

## Data Structure

-   Real data mostly consists of 4 types

    -   Character

    -   Logical

    -   Integer

    -   Numeric

## Character Data

-   Text information

    -   Called strings in Python

-   Contained in "" or ''

```{r}
#| echo: true
#| eval: false

fav_color <- "black"

print(fav_color)

```

## Logical data

-   Also know as binary data, either TRUE or FALSE

-   TRUE and FALSE are not contained within quotes

```{r}
#| echo: true
#| eval: false

x <- TRUE

print(x)


# Used for conditionals

if (x) {
  print("Yes")
}
```

## Integer and Numeric Data

-   Numeric data is all real numbers, can be decimal or not

-   Integers are whole real numbers

```{r}
#| echo: true
#| eval: false

x <- 3.4
y <- 4
z <- 4L

print(class(x))
print(class(y))
print(class(z))
print(z)

x <- as.integer(x)

print(class(x))
print(x)
```

## R Packages

- To use a package
  - Install (Once)
  - Load (Every time)

- Tidyverse is widely used 

```{r}
#install.packages("tidyverse")
library(tidyverse)
```


## Local Data

-   Data can be imputed directly into R

-   Tables can be useful for displaying data

```{r}
#| echo: true
#| eval: false

# Creates the numerical data in the table
tab <- as.table(rbind(c(762, 327, 468, 593), c(484, 239, 477, 530)))

# Labels the rows then the columns 
dimnames(tab) <- list(c("Honors", "Not Honors"), Grade = c("ninth", "tenth", "eleventh", "twelfth"))
```

## Local Data (Image)

```{r}
# Since it's a new slide the table has to be remade before being rendered
tab <- as.table(rbind(c(762, 327, 468, 593), c(484, 239, 477, 530)))
dimnames(tab) <- list(c("Honors", "Not Honors"), Grade = c("ninth", "tenth", "eleventh", "twelfth"))

tab
```
## Manipulating Local Data

```{r}
#| echo: true

# Only displays the first 2 columns
tab2 <- tab[, 1:2]
tab2

# Only displays the first row
tab3 <- tab[1,]
tab3

# Creates a new column avenging the grades by row
tab4 <- cbind(tab, Average = rowMeans(tab))
tab4
```

## Visualizing Local Data

- A bar graph is the best option to visualize this data

```{r}
#| echo: true
#| eval: false

# A simple bar graph
barplot(tab)

# Turning the table into a data frame
df_tab <- as.data.frame(tab)

# Tidyverse subsection ggplots are cleaner
ggplot(df_tab, aes(x = Grade, y = Freq, fill = Var1)) +
  geom_bar(stat = "identity", position = "fill")
```

## Visualizing Local Data (Image)

```{r}
barplot(tab)
```

## Visualizing Local Data (Another Image)

```{r}
df_tab <- as.data.frame(tab)
ggplot(df_tab, aes(x = Grade, y = Freq, fill = Var1)) +
  geom_bar(stat = "identity", position = "fill")
```

## Prexisting Data

- Data file needs to be in your current repository to be 
accessed

- Use setwd("path/to/repository") to change your current repository

- Use getwd() to determine your current repository

```{r}
#| echo: true
# A csv file can be imported via read.csv
player_stats <- read.csv("nba_player_stats_24-25.csv")
```

## Finding a Dataset

 - Countless datasets are available for download online
 
 - Quality statistical analysis is dependent on quality data
 
 - The data we will be using is from the website Kaggle, found [here](https://www.kaggle.com/datasets/eduardopalmieri/nba-player-stats-season-2425)

## First Look at NBA Data

- Displays the stats of every player for every game

- One observation (row) is one player's personal record for a 
specific game

- Variable descriptions can be found 
[here](https://www.kaggle.com/datasets/eduardopalmieri/nba-player-stats-season-2425)

```{r}
#| echo: true
#| eval: false
# str() gives a helpful overview
str(player_stats)

# head() shows the first few rows
head(player_stats)

# Quick scatterplot to look for irregularities in the dating
ggplot(player_stats, aes(Data, PTS)) +
  geom_point() +
  scale_x_discrete(breaks = range(player_stats$Data))
```

## First Look at NBA Data (Image)

```{r}
# The expand command isn't relevant so I didn't show it in the last slide, but it gives the graph some breathing room on the side
ggplot(player_stats, aes(Data, PTS)) +
  geom_point() +
  scale_x_discrete(breaks = range(player_stats$Data), expand = expansion(mult = c(0, 0.05)))
```


## NBA Data Clean Up 

- We prefer variable names to be all lowercase

- The Res variable should be logical


```{r}
#| echo: true
# Turns variable names lowercase
names(player_stats) <- tolower(names(player_stats))

# Changes res variable into logical win variable
player_stats <- player_stats %>%
  mutate(win = res == "W") %>%
  select(-res)
```

## Manipulating NBA Data

- What if we only want to look at the Lakers players?
  - How many points have they scored so far?

- What if we want to find the average statistics for each 
player?

```{r}
#| echo: true
# Filtering the data to only show the Lakers players
la_players <- player_stats %>%
  filter(tm == "LAL")

# Adding up total points scored variable
la_points <- sum(la_players$pts)
print(la_points)

# group_by function combines rows, summarise creates variables based off the previous dataset's variables
player_avg <- player_stats %>%
  group_by(player) %>%
  summarise(tm = first(tm), avg_pts = mean(pts), avg_mp = mean(mp), avg_trb = mean(trb), avg_ast = mean(ast))
```

## Visualizing NBA Data

- Using the la_players data frame
  - Visualize total points scored per game
  - Graph total points over time, differentiate each player

- Using the player_avg
  - Visualize average points per game
  - Graph average points per game by average minutes played,
  differentiate each team
  
## Visualizing LA Data

- Visualizing total points scored per game
  - One variable, numerical data -> histogram works well

```{r}
#| echo: true
#| eval: false
# Creating histogram, binwidth sets the length of the bars
ggplot(la_players, aes(pts)) +
  geom_histogram(binwidth = 5)
```

- Graphing total points over time, differentiating each player
  - Three variables, two numerical, one categorical
  - Scatterplot with extra variable linked to point color

```{r}
#| echo: true
#| eval: false
# Creating scatterplot, position_jitter spaces data
ggplot(la_players, aes(data, pts, color = player)) +
  geom_point(position = position_jitter(width = 0.2))
```

## Visualizing LA Data (Image)

```{r}
ggplot(la_players, aes(pts)) +
  geom_histogram(binwidth = 5)
```

## Histogram Questions

- What happens if we change the binwidth?

- Can we swap binwidth out with anything?

- What other variables can we make histograms with?

## Visualizing LA Data (Another Image)

```{r}
ggplot(la_players, aes(data, pts, color = player)) +
  geom_point(position = position_jitter(width = 0.2))
```

## Scatterplot Questions

- What happens if we take out position_jitter?

- Can we focus on just a portion of the data?
  - Hint: filter()
  
- How can we clean up the x axis ticks?

## Visualizing Player Averages Data

- Visualizing average points per game
  - One variable, numerical data -> histogram works well
  
```{r}
#| echo: true
#| eval: false
# Creating histogram, binwidth sets the length of the bars
ggplot(player_avg, aes(avg_pts)) +
  geom_histogram(binwidth = 2.5)
```  

- Graphing average points per game by average minutes played,
  while differentiating between teams
  - Three variables, two numerical, one categorical
  - Scatterplot with extra variable linked to point color

```{r}
#| echo: true
#| eval: false
# Creating scatterplot, shape linked to team
ggplot(player_avg, aes(avg_mp, avg_pts, color = tm)) +
  geom_point()
```

## Visualizing Player Averages Data (Image)

```{r}
# Creating histogram, binwidth sets the length of the bars
ggplot(player_avg, aes(avg_pts)) +
  geom_histogram(binwidth = 2.5)
```

## Histogram Questions

- Does this binwidth work well for the data?

- Would this look cleaner with a specific amount of bins?

- What variables in this data set might have a similar distribution?

## Visualizing Player Averages Data (Another Image)

```{r}
# Creating scatterplot, shape linked to team
ggplot(player_avg, aes(avg_mp, avg_pts, color = tm)) +
  geom_point()
```

## Scatterplot Questions

- How can we add a trend line to this graph?

- What portion of this data would it be helpful to focus on?
  - How can we do that?
  
- How can we clean up the graph labels?

## Thank You!

- Any remaining questions?

- Email:
  - tyler.adams@uconn.edu



