---
title: "Intro to R"
subtitle: "CSAS 2026"
author: "Tyler Adams"
format:
    revealjs:
        slide-number: true
        preview-links: true
        theme: default
---

## Prerequisites

- A laptop with [R](https://cran.r-project.org/)
and [Positron](https://positron.posit.co/download.html)
/[RStudio](https://posit.co/download/rstudio-desktop/) installed

- No experience needed

## Outline

- R basics

- R Syntax

- Data structure

- Practical R uses

## Why R?

- Runs on all major operating systems

- Open source

- Intuitive syntax

- User-created packages constantly provide new features

## Positron and RStudio

- Both Integrated Development Environments (IDEs) created
by posit

- RSudio is older, Positron released very recently

- Both have 4 main sections
  - Code editor
  - Variables
  - Console
  - Plots

## R syntax

- Type code into console

```{r}
#| echo: true
# Example 1
print("Hello world")
```
```{r}
#| echo: true
# Example 2
x <- 2
y = 3
```

- Both "=" and "<-" can be used to assign values to variables

- Look at the variables window

## R syntax, continued

- Basic operation signs

```{r}
#| echo: true
# Addition
print(2 + 2)

# Subtraction
print(4 - 2)

# Multiplication (variables can be used too)
print(y * x)

# Division
print(x / 3)

# Exponentiation
print(x^2)
# or
print(x**2)
```

## Data Structure

- Real data mostly consists of 4 types

  - Logical
  
  - Numeric
  
  - Integer
  
  - Character
